Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-17 doc/MS_v3.Rmd
sorta almost complete intro draft

Although meiotic recombination is required for successful gametogenesis in most species that reproduce sexually, the rate of crossing over varies among individuals. The number of crossovers per meiotic cell is the  genome-wide recombination rate. Differences in recombination rate between females and males are perhaps the most striking form of this variation. Existing data fail to directly address the extent to which recombination experiences similar evolutionary pressures. To fill this gap, we measured meiotic recombination in both sexes for a panel of house mouse cross three subspecies. Using inbred strains and single cell immunohistochemistry allowed us to place sex-specific observations within the same genetic background and meiotic context. Our results indicate highly discordant evolutionary patterns in the two sexes. Whereas male recombination rates show evidence of rapid evolution over short timescales, female recombination rates measured in the same strains are mostly static. These results strongly indicate that house mouse has two, one for each sex, genome wide recombination rates which display distinct evolutionary trajectories.

@@

  Meiosis can be reduced to the expression (2n -> 4n -> 2n -> 1n) which reflects the process of a diploid genome producing haploid gametes. In most species the meiotic program relies on crossovers to ensure the correct separation of chromosomes at the reductional division (4n to 2n) [@CahoonLibuda2019]. The total number of COs at this stage of meiosis is equal to the genome wide recombination rate (gwRR). In addition to this functional purpose, the gwRR can regulate a populations’ response to selection, and determine the fate of novel mutations by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds (Charlesworth 1993 cite).  This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation [@begun1992; @nachman2012].

<!-- ### 1. Understanding levels of variation in gwRR -->

  **Due to it's role in facilitating chromosomes disjunction, the genome wide recombination rate can be thought of as a cell-based metric which is tightly connected to organisms fitness and fertility.**  Ensuring at least one crossover per chrm while minimizing the rates of DNA damage and ectopic recombination are thought to regulate the lower and upper thresholds for a genome-wide recombiantion rate [@nagaoka2012] (Inoue). Yet, within these thresholds the genome wide recombination rate still varies across species and individuals. Sex is one of the most notable axes of this form of variation.

<!-- (examples, Humans 1.7 female to male
alligators female to male ratio is X, temperature dependent sex determination) -->

<!-- The number of haploid pairs of chromosomes is strong predictor for genome wide recombination rate across large divergence scales [@ottoPaysuer2019; @stapley_variation_2017]. -->

<!-- 2.Gap is due to Lack of comparable measures -->

An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog,  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep?) sheep [@johnston2016_soay], and mouse (CC cite) in which the role of sex is confounded with the contributions of genetic variation.

<!-- 3.	Open Q, no first principle -->

Even under anisogamy, where gametogenesis is modified to produce distinct gametes across sexes, the meiotic program is the same. There are no first principles which would predict the evolution of sexually dimorphic recombination rates. Yet heterochiasmy is commonly observed in dioecious species, suggesting that other meiotic traits which distinguish the gametes and their meiotic programs, for example symmetrical vs asymmetrical cell division, may impose selection for sexually dimorphic recombination rates. Although it is clear that the relationship between female and male recombination rates can differ among species, comparisons between and within closely related species are missing. Direct contrasts between females and males across a common, diverse set of genomic backgrounds would reveal whether the recombination rate evolves differently in the sexes.

<!-- 4.	Additional metrics (more nuanced) -->  
  Heterochiasimy was defined as sexual dimorphism in the total counts measures of recombination (Lenormand 2003). It's important to note that the genome-wide recombiantion rate can be decomposed into single chromosome patterns, refered to as recombination landscapes. Recently two specific features have been highlighted as conserved patterns related heterochiasmy, specifically the recombination landscape. In a review of data from mouse, celegans and worms with meiotic chromosome morphilogy (axis-loop structure) quantified for both sexes, @CahoonLibuda2019 find that the sex with longer chromosome axis generally has higher recombination rates. A survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement [@sardell_sex_2020]. These observations raise the question of how conservation in sexually dimorphic patterns are maintained with the gwRR and it’s decomposed traits.

-----------

<!-- ### 2.  The House Mouse is a great model for uncovering evolutionary patterns at a short timescale -->

  The house mouse species, (_Mus musculus_), complex arose from a recent radiation (0.5 mya to 5 mya) (@geraldes2011; @geraldes2008) providing an opportunity to interrogate natural variation at short evolutionary scales. This feature narrows the mutational space for identifying variants which would cause gwRR variation.  Classical lab strains of mice have generated extensive knowledge central to meiosis and outcomes on recombination through extensive studies on the genes involved in the meiotic recombination pathway **(cite)** and previous crosses for understanding the genetic architecture of recombination rate variation [@dumont2011; @murdoch2010; @Wang2017island]. Additionally due to their global distribution allows us to compare samples of natural diversity from a broad geographic range.

  **Wild derived inbred strains generate the best comparison of females and males, besides the sex chromosomes, the mouse for each genome are almost identical**. Inbred strains enable one of the most direct comparison of both female and male gametogenesis and enable us to conclude that sex is a primary factor.

  House mouse is well suited for single cell cytology approaches (due to large cell sizes). The single cell level allows integration of data at a closer connection to the molecular pathway and meiotic program. (Allows us to compare counts from cells at is the same meiotic (molecular pathway context)).

@@

---------------

  (we quantify gwRR from both sexes, using house mouse inbred strains as a common genetic background). We observed highly distinct evolutionary trajectories between males and females. This is due to rapid male specific evolution in the gwRR from two strains in musculus and 1 strain in molossinus. (this pattern of rapid evolution was not observed in females). We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale.

<!-- we decompsed the gwRR into single chromosome features to access data which  reflects the non-random processes of crossover formation on individual chromosomes.-->

  The recombination landscape for single chromosome scale can be general summarized by i) spacing of crossovers and ii) the linear area of synapsed homologous chromosomes where crossovers can form. We quantified the general recombination landscape because patterns in the recombination landscape are important to document because they can shed light on constraints and regulation of the processes ( could end up leading to variation in the genome wide recombination rates. Our results confirm (well established) previously reported patterns for sex differences in the general recombination landscape – and add by demonstrating that these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns. Conserved patterns of the recombination landscape of the single chromosome level – reflect constraints  At the single chromosome level – (crossover number is conserved – suggesting processes regulating the rec landscape at this scale might face similar constraints

>>>>

Although meiotic recombination is required for successful gametogenesis in most species that reproduce sexually, the rate of crossing over varies among individuals. The number of crossovers per meiotic cell is the  genome-wide recombination rate. Differences in recombination rate between females and males are perhaps the most striking form of this variation. Existing data fail to directly address the extent to which recombination experiences similar evolutionary pressures. To fill this gap, we measured meiotic recombination in both sexes for a panel of house mouse cross three subspecies. Using inbred strains and single cell immunohistochemistry allowed us to place sex-specific observations within the same genetic background and meiotic context. Our results indicate highly discordant evolutionary patterns in the two sexes. Whereas male recombination rates show evidence of rapid evolution over short timescales, female recombination rates measured in the same strains are mostly static. These results strongly indicate that house mouse has two, one for each sex, genome-wide recombination rates which display distinct evolutionary trajectories.

@@

  Meiosis can be reduced to the expression (2n -> 4n -> 2n -> 1n) which reflects the process of a diploid genome producing haploid gametes.  The total number of COs at this stage of meiosis is equal to the genome-wide recombination rate (gwRR). Meiosis (mixing of haploid gametes) is an important process for evolution.  Meiotic recombination can regulate a populations’ response to selection, and determine the fate of novel mutations by transferring beneficial mutations onto novel genetic backgrounds or by breaking linkage of negative mutations from beneficial genetic backgrounds (Charlesworth 1993 cite).  This process shapes the genomic patterns of genetic variation, with high recombination areas of genome having more nucleotide variation while areas of low recombination have decreased genetic variation [@begun1992; @nachman2012]. At the individual level (fertility of a individual), the meiotic program relies on crossovers (the products of meiotic recombination), to ensure the correct separation of chromosomes at the reductional division (4n to 2n) [@CahoonLibuda2019].

<!-- In most species the meiotic program relies on crossovers to ensure the correct separation of chromosomes at the reductional division (4n to 2n) [@CahoonLibuda2019].  ### 1. Understanding levels of variation in gwRR -->

  **Due to it's role in facilitating chromosomes disjunction, the genome-wide recombination rate can be thought of as a cell-based metric which is tightly connected to organisms fitness and fertility.**  (thus clear target of evolution). Ensuring at least one crossover per chrm while minimizing the rates of DNA damage and ectopic recombination are thought to regulate the lower and upper thresholds for a genome-wide recombiantion rate [@nagaoka2012] (Inoue). Yet, within these thresholds the genome-wide recombination rate still varies across species and individuals. Sex is one of the most notable axes of this form of variation. 
  There are two classes of sexual dimorphism in recombiantion rates; the first is called achiasmy and refered to where one sex recombines (Bell 1982 – Huxley 1928, Lenormand Duthiel 2005). The second, which is the focus of this manuscript, is the instance where both sexes recombine (Lenormand and Duthiel 2003). An understanding of how sex shapes the evolution of recombination cannot be achieved with available data. Comprehensive comparisons of female and male recombination rates usually come from outbred populations humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], dog [@wong2010dog],  cattle [@ma2015_cattle; @Shen2018_cattle], (other sheep?) sheep [@johnston2016_soay], and mouse [@srivastava2017CC] in which the role of sex is confounded with the contributions of genetic variation.

<!-- 3.	Open Q, no first principle -->  
In anisogamy ist systems where gametogenesis is modified to produce distinct gametes across sexes, the meiotic program is the same (gorelick?). There are no first principles which would predict the evolution of sexually dimorphic genome-wide recombination rates. Yet heterochiasmy is commonly observed in dioecious species, suggesting that other meiotic traits which distinguish the gametes and their meiotic programs, for example symmetrical vs asymmetrical cell division, may impose selection for sexually dimorphic recombination rates. Although it is clear that the relationship between female and male recombination rates can differ among species, comparisons between and within closely related species are missing. Direct contrasts between females and males across a common, diverse set of genomic backgrounds would reveal whether the recombination rate evolves differently in the sexes.

<!-- 4.	Additional metrics (more nuanced) -->

  The genome-wide recombiantion rate can be decomposed into single chromosome patterns, refered to as recombination landscapes. **(at this scale –there are also relevant features of heterochiasmy).** Recently two specific features have been highlighted as conserved patterns related heterochiasmy, specifically the recombination landscape. In a review of data from mouse, celegans and worms with meiotic chromosome morphilogy (axis-loop structure) quantified for both sexes, @CahoonLibuda2019 find that the sex with longer chromosome axis generally has higher recombination rates. A survey of 51 species shows there are conserved broad scale recombination landscape differences. Generally males have telomere-bias crossover placement and females have more uniform placement [@sardell_sex_2020]. These observations raise the question of how conservation in sexually dimorphic patterns are maintained with the gwRR and it’s decomposed traits.

-----------
<!-- ### 2.  The House Mouse is a great model for uncovering evolutionary patterns at a short timescale -->

  The house mouse species, (_Mus musculus_), complex arose from a recent radiation (0.5 mya to 5 mya) (@geraldes2011; @geraldes2008) providing an opportunity to interrogate natural variation at short evolutionary scales. This feature narrows the mutational space for identifying variants which lead to genome-wide recombination rate variation.  Classical lab strains of mice have generated extensive knowledge central to meiosis and outcomes on recombination through extensive studies on the genes involved in the meiotic recombination pathway **(Shementi? Hassold? cite)** and previous crosses for understanding the genetic architecture of recombination rate variation [@dumont2011; @murdoch2010; @Wang2017island]. Additionally due to their global distribution allows comparisons of from a broad geographic range that might be missed with classical inbred strains. Wild derived inbred strains of house mouse, enable one of the most direct comparison of female and male gametogenesis, since besides the sex chromosomes, the mouse for each genome are identical. Additionally, House mouse is well suited for single cell cytology approaches, enabling integration of data at a closer connection to the molecular pathway and meiotic program.

@@

  In this study / paper, We quantify gwRRs from both sexes, in a panel of inbred mouse strains. We observed highly distinct evolutionary trajectories between males and females. This is due to rapid male specific evolution in the gwRR from two strains in musculus and 1 strain in molossinus. We observed up to a 30% difference (translating into approximately ~7 more crossovers per cell), a surprising amount considering the short evolutionary time scale. This pattern of rapid evolution was not observed in females.

<!-- we decompsed the gwRR into single chromosome features to access data which  reflects the non-random processes of crossover formation on individual chromosomes.  The recombination landscape for single chromosome scale can be summarized by i) spacing of crossovers and ii) the linear area of synapsed homologous chromosomes where crossovers can form. -->

  In our comparisons of recombiantion landscapes across sex, our results confirm previously reported patterns for sex differences in the general recombination landscape [@sardell_sex_2020]. Intriguingly,  these patterns hold even in the case of genome more crossovers, but general landscape patterns still maintain sex-specific patterns.

  Our comparison of recombination landscapes framed within the rapid male specific evolution of gwRR revealed unexpected patterns - <**put something interesting here**>.
